\hypertarget{nim-standard-library}{%
\section{Nim Standard Library}\label{nim-standard-library}}

\begin{description}
\item[Author]
Andreas Rumpf
\item[Version]
1.3.5
\end{description}

Nim's library is divided into \emph{pure libraries}, \emph{impure
libraries} and \emph{wrappers}.

Pure libraries do not depend on any external \texttt{*.dll} or
\texttt{lib*.so} binary while impure libraries do. A wrapper is an
impure library that is a very low-level interface to a C library.

Read this \href{apis.html}{document} for a quick overview of the API
design.

\hypertarget{nimble}{%
\subsection{Nimble}\label{nimble}}

Nim's standard library only covers the basics, check out
\url{https://nimble.directory/} for a list of 3rd party packages.

\hypertarget{pure-libraries}{%
\subsection{Pure libraries}\label{pure-libraries}}

\hypertarget{automatic-imports}{%
\subsubsection{Automatic imports}\label{automatic-imports}}

\begin{itemize}
\tightlist
\item
  \href{system.html}{system} Basic procs and operators that every
  program needs. It also provides IO facilities for reading and writing
  text and binary files. It is imported implicitly by the compiler. Do
  not import it directly. It relies on compiler magic to work.
\item
  \href{threads.html}{threads} Basic Nim thread support. \textbf{Note}:
  This is part of the system module. Do not import it explicitly.
  Enabled with \texttt{-\/-threads:on}.
\item
  \href{channels.html}{channels} Nim message passing support for
  threads. \textbf{Note}: This is part of the system module. Do not
  import it explicitly. Enabled with \texttt{-\/-threads:on}.
\end{itemize}

\hypertarget{core}{%
\subsubsection{Core}\label{core}}

\begin{itemize}
\tightlist
\item
  \href{bitops.html}{bitops} Provides a series of low level methods for
  bit manipulation.
\item
  \href{cpuinfo.html}{cpuinfo} This module implements procs to determine
  the number of CPUs / cores.
\item
  \href{endians.html}{endians} This module contains helpers that deal
  with different byte orders.
\item
  \href{lenientops.html}{lenientops} Provides binary operators for mixed
  integer/float expressions for convenience.
\item
  \href{locks.html}{locks} Locks and condition variables for Nim.
\item
  \href{macros.html}{macros} Contains the AST API and documentation of
  Nim for writing macros.
\item
  \href{rlocks.html}{rlocks} Reentrant locks for Nim.
\item
  \href{typeinfo.html}{typeinfo} Provides (unsafe) access to Nim's run
  time type information.
\item
  \href{typetraits.html}{typetraits} This module defines compile-time
  reflection procs for working with types.
\item
  \href{volatile.html}{volatile} This module contains code for
  generating volatile loads and stores, which are useful in embedded and
  systems programming.
\end{itemize}

\hypertarget{algorithms}{%
\subsubsection{Algorithms}\label{algorithms}}

\begin{itemize}
\tightlist
\item
  \href{algorithm.html}{algorithm} Implements some common generic
  algorithms like sort or binary search.
\item
  \href{sequtils.html}{sequtils} This module implements operations for
  the built-in seq type which were inspired by functional programming
  languages.
\end{itemize}

\hypertarget{collections}{%
\subsubsection{Collections}\label{collections}}

\begin{itemize}
\tightlist
\item
  \href{critbits.html}{critbits} This module implements a \emph{crit bit
  tree} which is an efficient container for a sorted set of strings, or
  for a sorted mapping of strings.
\item
  \href{deques.html}{deques} Implementation of a double-ended queue. The
  underlying implementation uses a \texttt{seq}.
\item
  \href{heapqueue.html}{heapqueue} Implementation of a heap data
  structure that can be used as a priority queue.
\item
  \href{intsets.html}{intsets} Efficient implementation of a set of ints
  as a sparse bit set.
\item
  \href{lists.html}{lists} Nim linked list support. Contains singly and
  doubly linked lists and circular lists ("rings").
\item
  \href{options.html}{options} The option type encapsulates an optional
  value.
\item
  \href{sets.html}{sets} Nim hash and bit set support.
\item
  \href{sharedlist.html}{sharedlist} Nim shared linked list support.
  Contains shared singly linked list.
\item
  \href{sharedtables.html}{sharedtables} Nim shared hash table support.
  Contains shared tables.
\item
  \href{tables.html}{tables} Nim hash table support. Contains tables,
  ordered tables and count tables.
\end{itemize}

\hypertarget{string-handling}{%
\subsubsection{String handling}\label{string-handling}}

\begin{itemize}
\tightlist
\item
  \href{cstrutils.html}{cstrutils} Utilities for \texttt{cstring}
  handling.
\item
  \href{editdistance.html}{std/editdistance} This module contains an
  algorithm to compute the edit distance between two Unicode strings.
\item
  \href{encodings.html}{encodings} Converts between different character
  encodings. On UNIX, this uses the \texttt{iconv} library, on Windows
  the Windows API.
\item
  \href{parseutils.html}{parseutils} This module contains helpers for
  parsing tokens, numbers, identifiers, etc.
\item
  \href{pegs.html}{pegs} This module contains procedures and operators
  for handling PEGs.
\item
  \href{punycode.html}{punycode} Implements a representation of Unicode
  with the limited ASCII character subset.
\item
  \href{ropes.html}{ropes} This module contains support for a
  \emph{rope} data type. Ropes can represent very long strings
  efficiently; especially concatenation is done in O(1) instead of O(n).
\item
  \href{strformat.html}{strformat} Macro based standard string
  interpolation / formatting. Inspired by Python's \texttt{f}-strings.
\item
  \href{strmisc.html}{strmisc} This module contains uncommon string
  handling operations that do not fit with the commonly used operations
  in strutils.
\item
  \href{strscans.html}{strscans} This module contains a \texttt{scanf}
  macro for convenient parsing of mini languages.
\item
  \href{strtabs.html}{strtabs} The \texttt{strtabs} module implements an
  efficient hash table that is a mapping from strings to strings.
  Supports a case-sensitive, case-insensitive and style-insensitive
  modes.
\item
  \href{strutils.html}{strutils} This module contains common string
  handling operations like changing case of a string, splitting a string
  into substrings, searching for substrings, replacing substrings.
\item
  \href{unicode.html}{unicode} This module provides support to handle
  the Unicode UTF-8 encoding.
\item
  \href{unidecode.html}{unidecode} It provides a single proc that does
  Unicode to ASCII transliterations. Based on Python's Unidecode module.
\item
  \href{wordwrap.html}{std/wordwrap} This module contains an algorithm
  to wordwrap a Unicode string.
\end{itemize}

\hypertarget{time-handling}{%
\subsubsection{Time handling}\label{time-handling}}

\begin{itemize}
\tightlist
\item
  \href{monotimes.html}{std/monotimes} The {monotimes} module implements
  monotonic timestamps.
\item
  \href{times.html}{times} The \texttt{times} module contains support
  for working with time.
\end{itemize}

\hypertarget{generic-operating-system-services}{%
\subsubsection{Generic Operating System
Services}\label{generic-operating-system-services}}

\begin{itemize}
\tightlist
\item
  \href{distros.html}{distros} This module implements the basics for OS
  distribution ("distro") detection and the OS's native package manager.
  Its primary purpose is to produce output for Nimble packages, but it
  also contains the widely used \textbf{Distribution} enum that is
  useful for writing platform specific code. See
  \href{packaging.html}{packaging} for hints on distributing Nim using
  OS packages.
\item
  \href{dynlib.html}{dynlib} This module implements the ability to
  access symbols from shared libraries.
\item
  \href{marshal.html}{marshal} Contains procs for serialization and
  deserialization of arbitrary Nim data structures.
\item
  \href{memfiles.html}{memfiles} This module provides support for memory
  mapped files (Posix's \texttt{mmap}) on the different operating
  systems.
\item
  \href{os.html}{os} Basic operating system facilities like retrieving
  environment variables, reading command line arguments, working with
  directories, running shell commands, etc.
\item
  \href{osproc.html}{osproc} Module for process communication beyond
  \texttt{os.execShellCmd}.
\item
  \href{streams.html}{streams} This module provides a stream interface
  and two implementations thereof: the {FileStream} and the
  {StringStream} which implement the stream interface for Nim file
  objects ({File}) and strings. Other modules may provide other
  implementations for this standard stream interface.
\item
  \href{terminal.html}{terminal} This module contains a few procedures
  to control the \emph{terminal} (also called \emph{console}). The
  implementation simply uses ANSI escape sequences and does not depend
  on any other module.
\end{itemize}

\hypertarget{math-libraries}{%
\subsubsection{Math libraries}\label{math-libraries}}

\begin{itemize}
\tightlist
\item
  \href{complex.html}{complex} This module implements complex numbers
  and their mathematical operations.
\item
  \href{fenv.html}{fenv} Floating-point environment. Handling of
  floating-point rounding and exceptions (overflow, zero-divide, etc.).
\item
  \href{math.html}{math} Mathematical operations like cosine, square
  root.
\item
  \href{mersenne.html}{mersenne} Mersenne twister random number
  generator.
\item
  \href{random.html}{random} Fast and tiny random number generator.
\item
  \href{rationals.html}{rationals} This module implements rational
  numbers and their mathematical operations.
\item
  \href{stats.html}{stats} Statistical analysis
\item
  \href{sums.html}{std/sums} Fast summation functions.
\end{itemize}

\hypertarget{internet-protocols-and-support}{%
\subsubsection{Internet Protocols and
Support}\label{internet-protocols-and-support}}

\begin{itemize}
\tightlist
\item
  \href{asyncdispatch.html}{asyncdispatch} This module implements an
  asynchronous dispatcher for IO operations.
\item
  \href{asyncfile.html}{asyncfile} This module implements asynchronous
  file reading and writing using \texttt{asyncdispatch}.
\item
  \href{asyncftpclient.html}{asyncftpclient} This module implements an
  asynchronous FTP client using the \texttt{asyncnet} module.
\item
  \href{asynchttpserver.html}{asynchttpserver} This module implements an
  asynchronous HTTP server using the \texttt{asyncnet} module.
\item
  \href{asyncnet.html}{asyncnet} This module implements asynchronous
  sockets based on the \texttt{asyncdispatch} module.
\item
  \href{asyncstreams.html}{asyncstreams} This module provides
  {FutureStream} - a future that acts as a queue.
\item
  \href{cgi.html}{cgi} This module implements helpers for CGI
  applications.
\item
  \href{cookies.html}{cookies} This module contains helper procs for
  parsing and generating cookies.
\item
  \href{httpclient.html}{httpclient} This module implements a simple
  HTTP client which supports both synchronous and asynchronous retrieval
  of web pages.
\item
  \href{mimetypes.html}{mimetypes} This module implements a mimetypes
  database.
\item
  \href{nativesockets.html}{nativesockets} This module implements a
  low-level sockets API.
\item
  \href{net.html}{net} This module implements a high-level sockets API.
  It replaces the \texttt{sockets} module.
\item
  \href{selectors.html}{selectors} This module implements a selector API
  with backends specific to each OS. Currently epoll on Linux and select
  on other operating systems.
\item
  \href{smtp.html}{smtp} This module implement a simple SMTP client.
\item
  \href{uri.html}{uri} This module provides functions for working with
  URIs.
\end{itemize}

\hypertarget{threading}{%
\subsubsection{Threading}\label{threading}}

\begin{itemize}
\tightlist
\item
  \href{threadpool.html}{threadpool} Implements Nim's
  \href{manual_experimental.html\#parallel-amp-spawn}{spawn}.
\end{itemize}

\hypertarget{parsers}{%
\subsubsection{Parsers}\label{parsers}}

\begin{itemize}
\tightlist
\item
  \href{htmlparser.html}{htmlparser} This module parses an HTML document
  and creates its XML tree representation.
\item
  \href{json.html}{json} High performance JSON parser.
\item
  \href{lexbase.html}{lexbase} This is a low level module that
  implements an extremely efficient buffering scheme for lexers and
  parsers. This is used by the diverse parsing modules.
\item
  \href{parsecfg.html}{parsecfg} The \texttt{parsecfg} module implements
  a high performance configuration file parser. The configuration file's
  syntax is similar to the Windows \texttt{.ini} format, but much more
  powerful, as it is not a line based parser. String literals, raw
  string literals and triple quote string literals are supported as in
  the Nim programming language.
\item
  \href{parsecsv.html}{parsecsv} The \texttt{parsecsv} module implements
  a simple high performance CSV parser.
\item
  \href{parseopt.html}{parseopt} The \texttt{parseopt} module implements
  a command line option parser.
\item
  \href{parsesql.html}{parsesql} The \texttt{parsesql} module implements
  a simple high performance SQL parser.
\item
  \href{parsexml.html}{parsexml} The \texttt{parsexml} module implements
  a simple high performance XML/HTML parser. The only encoding that is
  supported is UTF-8. The parser has been designed to be somewhat error
  correcting, so that even some "wild HTML" found on the Web can be
  parsed with it.
\end{itemize}

\hypertarget{docutils}{%
\subsubsection{Docutils}\label{docutils}}

\begin{itemize}
\tightlist
\item
  \href{highlite.html}{packages/docutils/highlite} Source highlighter
  for programming or markup languages. Currently only few languages are
  supported, other languages may be added. The interface supports one
  language nested in another.
\item
  \href{rst.html}{packages/docutils/rst} This module implements a
  reStructuredText parser. A large subset is implemented. Some features
  of the markdown wiki syntax are also supported.
\item
  \href{rstast.html}{packages/docutils/rstast} This module implements an
  AST for the reStructuredText parser.
\item
  \href{rstgen.html}{packages/docutils/rstgen} This module implements a
  generator of HTML/Latex from reStructuredText.
\end{itemize}

\hypertarget{xml-processing}{%
\subsubsection{XML Processing}\label{xml-processing}}

\begin{itemize}
\tightlist
\item
  \href{xmltree.html}{xmltree} A simple XML tree. More efficient and
  simpler than the DOM. It also contains a macro for XML/HTML code
  generation.
\item
  \href{xmlparser.html}{xmlparser} This module parses an XML document
  and creates its XML tree representation.
\end{itemize}

\hypertarget{generators}{%
\subsubsection{Generators}\label{generators}}

\begin{itemize}
\tightlist
\item
  \href{htmlgen.html}{htmlgen} This module implements a simple XML and
  HTML code generator. Each commonly used HTML tag has a corresponding
  macro that generates a string with its HTML representation.
\end{itemize}

\hypertarget{hashing}{%
\subsubsection{Hashing}\label{hashing}}

\begin{itemize}
\tightlist
\item
  \href{base64.html}{base64} This module implements a base64 encoder and
  decoder.
\item
  \href{hashes.html}{hashes} This module implements efficient
  computations of hash values for diverse Nim types.
\item
  \href{md5.html}{md5} This module implements the MD5 checksum
  algorithm.
\item
  \href{oids.html}{oids} An OID is a global ID that consists of a
  timestamp, a unique counter and a random value. This combination
  should suffice to produce a globally distributed unique ID. This
  implementation was extracted from the Mongodb interface and it thus
  binary compatible with a Mongo OID.
\item
  \href{sha1.html}{std/sha1} This module implements a sha1 encoder and
  decoder.
\end{itemize}

\hypertarget{miscellaneous}{%
\subsubsection{Miscellaneous}\label{miscellaneous}}

\begin{itemize}
\tightlist
\item
  \href{browsers.html}{browsers} This module implements procs for
  opening URLs with the user's default browser.
\item
  \href{colors.html}{colors} This module implements color handling for
  Nim.
\item
  \href{coro.html}{coro} This module implements experimental coroutines
  in Nim.
\item
  \href{logging.html}{logging} This module implements a simple logger.
\item
  \href{segfaults.html}{segfaults} Turns access violations or segfaults
  into a \texttt{NilAccessDefect} exception.
\item
  \href{sugar.html}{sugar} This module implements nice syntactic sugar
  based on Nim's macro system.
\item
  \href{unittest.html}{unittest} Implements a Unit testing DSL.
\item
  \href{varints.html}{std/varints} Decode variable length integers that
  are compatible with SQLite.
\end{itemize}

\hypertarget{modules-for-js-backend}{%
\subsubsection{Modules for JS backend}\label{modules-for-js-backend}}

\begin{itemize}
\tightlist
\item
  \href{asyncjs.html}{asyncjs} Types and macros for writing asynchronous
  procedures in JavaScript.
\item
  \href{dom.html}{dom} Declaration of the Document Object Model for the
  JS backend.
\item
  \href{jsconsole.html}{jsconsole} Wrapper for the \texttt{console}
  object.
\item
  \href{jscore.html}{jscore} Wrapper of core JavaScript functions. For
  most purposes you should be using the \texttt{math}, \texttt{json},
  and \texttt{times} stdlib modules instead of this module.
\item
  \href{jsffi.html}{jsffi} Types and macros for easier interaction with
  JavaScript.
\end{itemize}

\hypertarget{impure-libraries}{%
\subsection{Impure libraries}\label{impure-libraries}}

\hypertarget{regular-expressions}{%
\subsubsection{Regular expressions}\label{regular-expressions}}

\begin{itemize}
\tightlist
\item
  \href{re.html}{re} This module contains procedures and operators for
  handling regular expressions. The current implementation uses PCRE.
\end{itemize}

\hypertarget{database-support}{%
\subsubsection{Database support}\label{database-support}}

\begin{itemize}
\tightlist
\item
  \href{db_postgres.html}{db\_postgres} A higher level PostgreSQL
  database wrapper. The same interface is implemented for other
  databases too.
\item
  \href{db_mysql.html}{db\_mysql} A higher level MySQL database wrapper.
  The same interface is implemented for other databases too.
\item
  \href{db_sqlite.html}{db\_sqlite} A higher level SQLite database
  wrapper. The same interface is implemented for other databases too.
\end{itemize}

\hypertarget{wrappers}{%
\subsection{Wrappers}\label{wrappers}}

The generated HTML for some of these wrappers is so huge that it is not
contained in the distribution. You can then find them on the website.

\hypertarget{windows-specific}{%
\subsubsection{Windows specific}\label{windows-specific}}

\begin{itemize}
\tightlist
\item
  \href{winlean.html}{winlean} Contains a wrapper for a small subset of
  the Win32 API.
\item
  \href{registry.html}{registry} Windows registry support.
\end{itemize}

\hypertarget{unix-specific}{%
\subsubsection{UNIX specific}\label{unix-specific}}

\begin{itemize}
\tightlist
\item
  \href{posix.html}{posix} Contains a wrapper for the POSIX standard.
\item
  \href{posix_utils.html}{posix\_utils} Contains helpers for the POSIX
  standard or specialized for Linux and BSDs.
\end{itemize}

\hypertarget{regular-expressions-1}{%
\subsubsection{Regular expressions}\label{regular-expressions-1}}

\begin{itemize}
\tightlist
\item
  \href{pcre.html}{pcre} Wrapper for the PCRE library.
\end{itemize}

\hypertarget{gui-libraries}{%
\subsubsection{GUI libraries}\label{gui-libraries}}

\begin{itemize}
\tightlist
\item
  \href{iup.html}{iup} Wrapper of the IUP GUI library.
\end{itemize}

\hypertarget{database-support-1}{%
\subsubsection{Database support}\label{database-support-1}}

\begin{itemize}
\tightlist
\item
  \href{postgres.html}{postgres} Contains a wrapper for the PostgreSQL
  API.
\item
  \href{mysql.html}{mysql} Contains a wrapper for the mySQL API.
\item
  \href{sqlite3.html}{sqlite3} Contains a wrapper for SQLite 3 API.
\item
  \href{odbcsql.html}{odbcsql} interface to the ODBC driver.
\end{itemize}

\hypertarget{network-programming-and-internet-protocols}{%
\subsubsection{Network Programming and Internet
Protocols}\label{network-programming-and-internet-protocols}}

\begin{itemize}
\tightlist
\item
  \href{openssl.html}{openssl} Wrapper for OpenSSL.
\end{itemize}
